---
title: 'Reproducible Research: Course Project 2'
author: "Alicia Rodriguez"
date: "3/19/2017"
output: 
  html_document:
    keep_md: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Health and Economic Consequences of Weather Events between 1950 and 2011

## Synopsis

This documents provides an analysis of the impact on public health and on economy across the United States, in the time span 1950-2011. The impact on public healt will be measured as the number of deaths, the number of people injured, and the combination of both during each weather event considered. The economic impact will be measured as the value in $ of the properties damages, the value in $ of crop damages, and the combination of both caused by the weather events considered. The impact will be considered along the whole time span (1950-2011) and along a narrower time span of the last 10 years (2001-2011). The goal is for the responsibles for prioritizing resources for each severe weather event to evaluate if they want to reduce any of the variables (deaths, injuries, property damage, or crop damage) isolated, or considering any combination of them. Useful information as the frequency and implicit damage of each severe event is also provided to complete the context of the study.

## Downloading, Reading and Processing the Raw Data

First, let's download the data and load the content using the option as.is in order to load the contents with no transformations:
```{r downloading_data}
  download.file("https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2", 
                destfile="./original_data.csv.bz2", method="curl")
```
```{r reading_data, cache=TRUE}
  original_data<-read.csv("./original_data.csv.bz2", as.is=T)
```

Let's take a look at the first rows, as well as the structure:
```{r inspecting_data, cache=TRUE}
  head(original_data)
  str(original_data)
```

A first inspection sugests some transformations that could be useful, mainly related to factors in the event type.
```{r data_preprocessing}
  original_data$EVTYPE<-factor(original_data$EVTYPE)
```

Regarding the columns representing the damages (property and crop), they need to be transformed so that they hold their whole value (base value and modifying exponen like Millions, Billions...). Before transforming the alphabetical characters used to signify magnitude for property and crop damages, let's take a look at the values they may take:

```{r magnitude_characters}
unique(original_data$PROPDMGEXP)
unique(original_data$CROPDMGEXP)
```

We see that there are more symbols besides the "K", "M", and "B" specified in the documentation. We will treat lowercase "k" and "m" as the uppercase cases. We will also consider that h and H are for hundreds. For the numerical symbols, let's take a look if the numbers could be the exponents (e.g., 2 corresponds to 10+E02, i.e., hundreds). 

```{r crop_exponent_checking}
tail(cbind(original_data$CROPDMG[grep("[0-9]",original_data$CROPDMGEXP)],
      as.numeric(original_data$CROPDMGEXP[grep("[0-9]",original_data$CROPDMGEXP)])),n=10)
head(cbind(original_data$PROPDMG[grep("[0-9]",original_data$PROPDMGEXP)],
      as.numeric(original_data$PROPDMGEXP[grep("[0-9]",original_data$PROPDMGEXP)])),n=10)
```

We see that for some values in the first column which are zero, the second column values are different from zero, which makes no sense with the hypothesis of the second column being exponents. Therefore, we discard the hypothesis of the numbers in the columns EXP being exponents and leave the base values unmodified (i.e., as the value in PROPDMG and CROPDMG).

For the rest, since there is no information in the documentation, we will leave them as the value in PROPDMG and CROPDMG. Since there is no info on how billions should be interpreted, the transformation will be to multiply by 1,000,000,000.

```{r magnitude_calculation}
original_data$PROPDMGFULL<-original_data$PROPDMG
original_data$PROPDMGFULL[original_data$PROPDMGEXP=="H"]<-original_data$PROPDMG[original_data$PROPDMGEXP=="H"]*100
original_data$PROPDMGFULL[original_data$PROPDMGEXP=="h"]<-original_data$PROPDMG[original_data$PROPDMGEXP=="h"]*100
original_data$PROPDMGFULL[original_data$PROPDMGEXP=="K"]<-original_data$PROPDMG[original_data$PROPDMGEXP=="K"]*1000
original_data$PROPDMGFULL[original_data$PROPDMGEXP=="k"]<-original_data$PROPDMG[original_data$PROPDMGEXP=="k"]*1000
original_data$PROPDMGFULL[original_data$PROPDMGEXP=="M"]<-original_data$PROPDMG[original_data$PROPDMGEXP=="M"]*1000000
original_data$PROPDMGFULL[original_data$PROPDMGEXP=="m"]<-original_data$PROPDMG[original_data$PROPDMGEXP=="m"]*1000000
original_data$PROPDMGFULL[original_data$PROPDMGEXP=="B"]<-original_data$PROPDMG[original_data$PROPDMGEXP=="B"]*1000000000
original_data$PROPDMGFULL[original_data$PROPDMGEXP=="b"]<-original_data$PROPDMG[original_data$PROPDMGEXP=="b"]*1000000000

original_data$CROPDMGFULL<-original_data$CROPDMG
original_data$CROPDMGFULL[original_data$CROPDMGEXP=="h"]<-original_data$CROPDMG[original_data$CROPDMGEXP=="h"]*100
original_data$CROPDMGFULL[original_data$CROPDMGEXP=="H"]<-original_data$CROPDMG[original_data$CROPDMGEXP=="H"]*100
original_data$CROPDMGFULL[original_data$CROPDMGEXP=="K"]<-original_data$CROPDMG[original_data$CROPDMGEXP=="K"]*1000
original_data$CROPDMGFULL[original_data$CROPDMGEXP=="k"]<-original_data$CROPDMG[original_data$CROPDMGEXP=="k"]*1000
original_data$CROPDMGFULL[original_data$CROPDMGEXP=="M"]<-original_data$CROPDMG[original_data$CROPDMGEXP=="M"]*1000000
original_data$CROPDMGFULL[original_data$CROPDMGEXP=="m"]<-original_data$CROPDMG[original_data$CROPDMGEXP=="m"]*1000000
original_data$CROPDMGFULL[original_data$CROPDMGEXP=="b"]<-original_data$CROPDMG[original_data$CROPDMGEXP=="b"]*1000000000
original_data$CROPDMGFULL[original_data$CROPDMGEXP=="B"]<-original_data$CROPDMG[original_data$CROPDMGEXP=="B"]*1000000000
```

Since no operations with dates are being considered, and the analysis will be across United States (not centered in each state), the rest of the fields are left unmodified.

## Exploring the Impact on Public Health

For inspecting the impact of weather events on public health, the analysis will be focused on fatalities and injuries. Both the cumulative amount of each attribute and the mean value will be analyzed. This way, we can have an intuition on the weather events which cause a high impact because of their severeness and those wich cause a high impact because of their frequency.
```{r creating_public_health_summary}
total_fatalities<-aggregate(x = original_data$FATALITIES, by=list(original_data$EVTYPE), FUN=sum)
total_injuries<-aggregate(x = original_data$INJURIES, by=list(original_data$EVTYPE), FUN=sum)
mean_fatalities<-aggregate(x = original_data$FATALITIES, by=list(original_data$EVTYPE), FUN=mean)
mean_injuries<-aggregate(x = original_data$INJURIES, by=list(original_data$EVTYPE), FUN=mean)

names(total_fatalities)<-c("EVTYPE","TOTAL_FATALITIES")
names(total_injuries)<-c("EVTYPE","TOTAL_INJURIES")
names(mean_fatalities)<-c("EVTYPE","MEAN_FATALITIES")
names(mean_injuries)<-c("EVTYPE","MEAN_INJURIES")

publicHealth<-merge(total_fatalities,total_injuries,by = "EVTYPE")
publicHealth<-merge(publicHealth, mean_fatalities,by = "EVTYPE")
publicHealth<-merge(publicHealth,mean_injuries,by = "EVTYPE")

publicHealth$TOTAL<-publicHealth$TOTAL_FATALITIES+publicHealth$TOTAL_INJURIES
publicHealth$MEAN<-publicHealth$MEAN_FATALITIES+publicHealth$MEAN_INJURIES
publicHealth<-publicHealth[order(publicHealth$TOTAL,decreasing = T),]
```

Let's have a look at the 10 first rows of data frame created with the publich health impact data, i.e., at the weather events with the highest impact and, thus, those we will focus on.
```{r exploring_similar_rows}
head(publicHealth,n=10)
```

We can observe that there are at least two events, THUNDERSTORM WIND and TSTM WIND, which represent the same weather event, but are labeled with different names. Let's see if there are more cases similar to this one with thunderstorm winds (with no other weather event in its name) and tornados, which also seem to suffer a similar problem, and are the leading wheather events.
```{r exploring_thunderstorm_rows}
publicHealth[grep("^(\\btorn\\w+)",publicHealth[,"EVTYPE"],ignore.case = T),]
publicHealth[grep("^((\\bthun\\w+)|(\\btstm))( )(wind|winds)$",publicHealth[,"EVTYPE"],ignore.case = T),]
```

We see that the tornado case is not critical, as the other occurrences do not accumulate enough fatalities or injuries as to greatly modify the results. However, there are many occurrences of thunderstorm wind events. Therefore, let's merge all of them.
```{r merging_similar_rows}
thunderstormRows<-grep("^((\\bthun\\w+)|(\\btstm))( )(wind|winds)$",
                       publicHealth[,"EVTYPE"],
                       ignore.case = T)
publicHealth$TOTAL_FATALITIES[thunderstormRows[1]]<-sum(publicHealth$TOTAL_FATALITIES[thunderstormRows])
publicHealth$TOTAL_INJURIES[thunderstormRows[1]]<-sum(publicHealth$TOTAL_INJURIES[thunderstormRows])
publicHealth$TOTAL[thunderstormRows[1]]<-sum(publicHealth$TOTAL[thunderstormRows])
publicHealth<-publicHealth[-thunderstormRows[2:length(thunderstormRows)],]
```

If we check now the thunderstorm wind occurrences, we have the following:
```{r thunderstorm_wind_occurences}
publicHealth[grep("^((\\bthun\\w+)|(\\btstm))( )(wind|winds)$", publicHealth[,"EVTYPE"], ignore.case = TRUE),]
```

Since the data available is from back 1950, and it can be seen that the data collected at that old dates are quite different (less detailed) than the current one, let's make a narrower analysis ranging the last 10 years of the dataset (2001-2011). We are going to make the same transformations as with the whole data set, explained above.
```{r analysis_10_last_years}
#Select data ranging from 2001-2011
lasttenyears_data<-original_data[as.Date(original_data$BGN_DATE, "%m/%d/%Y %H:%M:%S")>as.Date("2001-01-01"),]

#Aggregate fatalities and injuries (total and mean)
total_fatalities_lasttenyears<-aggregate(x = lasttenyears_data$FATALITIES,
                                   by=list(lasttenyears_data$EVTYPE), FUN=sum)
mean_fatalities_lasttenyears<-aggregate(x = lasttenyears_data$FATALITIES,
                                   by=list(lasttenyears_data$EVTYPE), FUN=mean)
total_injuries_lasttenyears<-aggregate(x = lasttenyears_data$INJURIES,
                                 by=list(lasttenyears_data$EVTYPE), FUN=sum)
mean_injuries_lasttenyears<-aggregate(x = lasttenyears_data$INJURIES,
                                 by=list(lasttenyears_data$EVTYPE), FUN=mean)

names(total_fatalities_lasttenyears)<-c("EVTYPE","TOTAL_FATALITIES")
names(total_injuries_lasttenyears)<-c("EVTYPE","TOTAL_INJURIES")
names(mean_fatalities_lasttenyears)<-c("EVTYPE","MEAN_FATALITIES")
names(mean_injuries_lasttenyears)<-c("EVTYPE","MEAN_INJURIES")

#Merge all data in a new dataframe
publicHealth_lasttenyears<-merge(total_fatalities_lasttenyears,
                                 mean_fatalities_lasttenyears,
                                 by = "EVTYPE")
publicHealth_lasttenyears<-merge(publicHealth_lasttenyears,
                                 total_injuries_lasttenyears,
                                 by = "EVTYPE")
publicHealth_lasttenyears<-merge(publicHealth_lasttenyears,
                                 mean_injuries_lasttenyears,
                                 by = "EVTYPE")
                                 
#Calculate total of fatalities and injuries
publicHealth_lasttenyears$TOTAL<-publicHealth_lasttenyears$TOTAL_FATALITIES+
  publicHealth_lasttenyears$TOTAL_INJURIES
publicHealth_lasttenyears$MEAN<-publicHealth_lasttenyears$MEAN_FATALITIES+
  publicHealth_lasttenyears$MEAN_INJURIES

#Sort by total impact
publicHealth_lasttenyears<-publicHealth_lasttenyears[order(publicHealth_lasttenyears$TOTAL,
                                                           decreasing = T),]

#Check thunderstorm duplicate names
publicHealth_lasttenyears[grep("^((\\bthun\\w+)|(\\btstm))( )(wind|winds)$",publicHealth_lasttenyears[,"EVTYPE"],ignore.case = T),]

#Merge thunderstorm winds rows
thunderstormRows_lasttenyears<-grep("^((\\bthun\\w+)|(\\btstm))( )(wind|winds)$", publicHealth_lasttenyears[,"EVTYPE"], ignore.case = TRUE)
publicHealth_lasttenyears$TOTAL_FATALITIES[thunderstormRows_lasttenyears[1]]<-sum(publicHealth_lasttenyears$TOTAL_FATALITIES[thunderstormRows_lasttenyears])
publicHealth_lasttenyears$TOTAL_INJURIES[thunderstormRows_lasttenyears[1]]<-sum(publicHealth_lasttenyears$TOTAL_INJURIES[thunderstormRows_lasttenyears])
publicHealth_lasttenyears$TOTAL[thunderstormRows_lasttenyears[1]]<-sum(publicHealth_lasttenyears$TOTAL[thunderstormRows_lasttenyears])

publicHealth_lasttenyears<-publicHealth_lasttenyears[-thunderstormRows_lasttenyears[2:length(thunderstormRows_lasttenyears)],]
publicHealth_lasttenyears[grep("^((\\bthun\\w+)|(\\btstm))( )(wind|winds)$", publicHealth_lasttenyears[,"EVTYPE"], ignore.case = T),]
```

In the Results section, the results from this analysis will be exposed.

## Exploring the Impact on Economy

For inspecting the impact of weather events on the economy, the analysis will be focused on damages in properties and crop. Both the cumulative amount of each attribute and the mean value will be analyzed. This way, we can have an intuition on the weather events which cause a high impact because of their severeness and those wich cause a high impact because of their frequency.
```{r creating_economy_summary}
total_propdmg<-aggregate(x = original_data$PROPDMGFULL, by=list(original_data$EVTYPE), FUN=sum)
total_cropdmg<-aggregate(x = original_data$CROPDMGFULL, by=list(original_data$EVTYPE), FUN=sum)
mean_propdmg<-aggregate(x = original_data$PROPDMGFULL, by=list(original_data$EVTYPE), FUN=mean)
mean_cropdmg<-aggregate(x = original_data$CROPDMGFULL, by=list(original_data$EVTYPE), FUN=mean)

names(total_propdmg)<-c("EVTYPE","TOTAL_PROPDMG")
names(total_cropdmg)<-c("EVTYPE","TOTAL_CROPDMG")
names(mean_propdmg)<-c("EVTYPE","MEAN_PROPDMG")
names(mean_cropdmg)<-c("EVTYPE","MEAN_CROPDMG")

economy<-merge(total_propdmg,total_cropdmg,by = "EVTYPE")
economy<-merge(economy, mean_propdmg,by = "EVTYPE")
economy<-merge(economy, mean_cropdmg,by = "EVTYPE")

economy$TOTAL<-economy$TOTAL_PROPDMG+economy$TOTAL_CROPDMG
economy$MEAN<-economy$MEAN_PROPDMG+economy$MEAN_CROPDMG
economy<-economy[order(economy$TOTAL,decreasing = T),]
```

Let's have a look at the 10 first rows of data frame created with the economy impact data, i.e., at the weather events with the highest impact and, thus, those we will focus on.
```{r check_economy_summary}
head(economy,n=10)
```

In this case, no name repetitions appear in these first ten elements, thus no merging will be applied.

As done with health impact, an additional analysis of a narrower time spam will be performed, again ranging the last 10 years of the dataset (2001-2011). We are going to make the same transformations as with the whole data set, explained above.
```{r creating_economy_summary_lasttenyears}
#Aggregate data using the lasttenyears_data data frame previously created 
#for analysis of impact on public health
total_propdmg_lasttenyears<-aggregate(x = lasttenyears_data$PROPDMGFULL,
                                   by=list(lasttenyears_data$EVTYPE), FUN=sum)
mean_propdmg_lasttenyears<-aggregate(x = lasttenyears_data$PROPDMGFULL,
                                   by=list(lasttenyears_data$EVTYPE), FUN=mean)
total_cropdmg_lasttenyears<-aggregate(x = lasttenyears_data$CROPDMGFULL,
                                 by=list(lasttenyears_data$EVTYPE), FUN=sum)
mean_cropdmg_lasttenyears<-aggregate(x = lasttenyears_data$CROPDMGFULL,
                                 by=list(lasttenyears_data$EVTYPE), FUN=mean)

names(total_propdmg_lasttenyears)<-c("EVTYPE","TOTAL_PROPDMG")
names(total_cropdmg_lasttenyears)<-c("EVTYPE","TOTAL_CROPDMG")
names(mean_propdmg_lasttenyears)<-c("EVTYPE","MEAN_PROPDMG")
names(mean_cropdmg_lasttenyears)<-c("EVTYPE","MEAN_CROPDMG")

#Create new data frame with all aggregates
economy_lasttenyears<-merge(total_propdmg_lasttenyears,
                                 total_cropdmg_lasttenyears,
                                 by = "EVTYPE")
economy_lasttenyears<-merge(economy_lasttenyears,
                                 mean_propdmg_lasttenyears,
                                 by = "EVTYPE")
economy_lasttenyears<-merge(economy_lasttenyears,
                                 mean_cropdmg_lasttenyears,
                                 by = "EVTYPE")
                   
#Calculate totals              
economy_lasttenyears$TOTAL<-economy_lasttenyears$TOTAL_PROPDMG+
  economy_lasttenyears$TOTAL_CROPDMG
economy_lasttenyears$MEAN<-economy_lasttenyears$MEAN_PROPDMG+
  economy_lasttenyears$MEAN_CROPDMG

#Sort by total damage
economy_lasttenyears<-economy_lasttenyears[order(economy_lasttenyears$TOTAL,
                                                           decreasing = T),]
```

## Results

### Weather Impact on Public Health

Let's take a look first at the ten weather events with the highest impact on public health, this is, with the highest number of fatalities and injuries (combined) for the whole time span, and the last 10 years:
```{r summary_public_health_impact}
publicHealth[1:10,"EVTYPE"]
publicHealth_lasttenyears[1:10,"EVTYPE"]
```

We can observe that the most harmful weather events in all the time span and in the last ten years are roughly the same: tornados, excessive heat, thunderstorm wind, lightning and heat remaining in the first 5 positions. Floods in the last years have become less impactful with respect to the rest of the events. In these last years the hurricanes/typhoons and wildfires have appeared as a new important events impacting publich health. 

```{r plotting_public_health_impact, fig.cap="*Weather impact on public health, measured as number of deaths and people injured during the event.*", fig.width=10, fig.height=8}
layout(matrix(c(1,2,3,4,5,3), 2, 3, byrow = TRUE), 
  	widths=c(4,4,1), heights=c(1,1))

par(las=2)
par(mar=c(11,4,2,1)+0.1)
barplot(t(publicHealth[1:10,c("TOTAL_FATALITIES","TOTAL_INJURIES")]),
        main="Total Number of People Affected (historic)",
        log="y",
        names.arg=publicHealth$EVTYPE[1:10],
        horiz=FALSE,
        cex.axis = 0.8)
barplot(t(publicHealth_lasttenyears[1:10,c("TOTAL_FATALITIES","TOTAL_INJURIES")]),
        main="Total Number of People Affected (2001-2011)",
        log="y",
        names.arg=publicHealth_lasttenyears$EVTYPE[1:10],
        horiz=FALSE,
        cex.axis = 0.8)

# setup for no margins on the legend
par(mar=c(0, 0, 0, 0))
plot.new()
legend('center','groups',c("Fatalities","Injuries"), fill = c("black","grey"),horiz = FALSE)

par(las=2)
par(mar=c(11,4,2,1)+0.1)
barplot(t(publicHealth[1:10,c("MEAN_FATALITIES","MEAN_INJURIES")]),
        main="Mean Number of People Affected (historic)",
        log="y",
        names.arg=publicHealth$EVTYPE[1:10],
        horiz=FALSE,
        cex.axis = 0.8)
barplot(t(publicHealth_lasttenyears[1:10,c("MEAN_FATALITIES","MEAN_INJURIES")]),
        main="Mean Number of People Affected (2001-2011)",
        log="y",
        names.arg=publicHealth_lasttenyears$EVTYPE[1:10],
        horiz=FALSE,
        cex.axis = 0.8)
```

The figures show this behaviour in the total impact. Moreover, it can be seen which events provoke more fatalities, which allows to assess these weather events thus reducing the moratility rates. 

Taking a look at the mean values, we can see that although tornados are the events provoking, in total terms, the highest amount of fatalities and injuries, it is excessive heat the event leading to the highest mean value. This seems to suggest that tornados frequency might be higher than excessive heat, although the last one causes more impact per event. Let's analyze the frequency of the events to confirm our hypothesis:
```{r events_frequency,fig.width=8, fig.height=10, fig.cap="*Frequency of the 20 most frequent weather events*"}
#Counting weather events for the whole time span
events_frequency<-as.data.frame(table(original_data$EVTYPE))
names(events_frequency)<-c("EVTYPE","FREQUENCY")

#Merging thunderstorm wind rows for the whole time span
thunderstormRows<-grep("^((\\bthun\\w+)|(\\btstm))( )(wind|winds)$", events_frequency[,"EVTYPE"], ignore.case = TRUE)
events_frequency$FREQUENCY[thunderstormRows[1]]<-sum(events_frequency$FREQUENCY[thunderstormRows])
events_frequency<-events_frequency[-thunderstormRows[2:length(thunderstormRows)],]

#Sorting for the whole time span
events_frequency<-events_frequency[order(events_frequency$FREQUENCY,decreasing = TRUE),]

#Plotting for the whole time span
layout(matrix(c(1,2), 2, 1, byrow = TRUE), 
  	widths=c(1), heights=c(1,1))
par(las=2)
par(mar=c(3,17,2,2)+0.1)

#Plotting frequencies for the whole time span
barplot(t(events_frequency$FREQUENCY[1:20]/1000),
        main="Weather Events Frequency (historic, in thousands)",
        log="x",
        names.arg=events_frequency$EVTYPE[1:20],
        horiz=TRUE,
        cex.axis = 0.8)

#Counting events for the last 10 years
events_frequency_lasttenyears<-as.data.frame(table(lasttenyears_data$EVTYPE))
names(events_frequency_lasttenyears)<-c("EVTYPE","FREQUENCY")

#Merging thunderstorm wind rows for the last 10 years
thunderstormRows_lasttenyears<-grep("^((\\bthun\\w+)|(\\btstm))( )(wind|winds)$", events_frequency_lasttenyears[,"EVTYPE"], ignore.case = TRUE)
events_frequency_lasttenyears$FREQUENCY[thunderstormRows_lasttenyears]<-sum(events_frequency_lasttenyears$FREQUENCY[thunderstormRows_lasttenyears])
events_frequency_lasttenyears<-events_frequency_lasttenyears[-thunderstormRows_lasttenyears[2:length(thunderstormRows_lasttenyears)],]

#Sorting for the last 10 years
events_frequency_lasttenyears<-events_frequency_lasttenyears[order(events_frequency_lasttenyears$FREQUENCY,decreasing = TRUE),]

#Plotting frequencies for the last 10 years
barplot(t(events_frequency_lasttenyears$FREQUENCY[1:20]/1000),
        main="Weather Events Frequency (2001-2011, in thousands)",
        log="x",
        names.arg=events_frequency_lasttenyears$EVTYPE[1:20],
        horiz=TRUE,
        cex.axis = 0.8)
```

These plots confirm the hypothesis: tornados are among the most frequent events, whereas heat and excessive heat do not appear among the 20 most frequent events. Thus, events of extreme heat are rare, but with a high impact. Same thing applies to hurricane/typhoon events. On the other hand, thunderstorm winds and lightning are very frequent, but the impact of these types of events is small.

### Weather Impact on Economy

Regarding the impact on the economy, let's first take a look at the top 10 weather events causing the highest damage (property and crop combined), for the whole time span, and the last 10 years:
```{r summary_economy_impact}
economy[1:10,"EVTYPE"]
economy_lasttenyears[1:10,"EVTYPE"]
```

In this case, the events causing the highest impact on economy are floods and hurricans/typhoons, followed by tornados, storm surge, and hail. This trend is the same both in the whole time span and in the last 10 years. Let's look at the figures to have a more precise idea on the property and crop damage, as well as the mean values.
```{r plotting_economy_impact, fig.cap="*Weather impact on economy, measured as the value in $ of the damages caused by the weather event to properties and crop.*", fig.width=10, fig.height=8}
layout(matrix(c(1,2,3,4,5,3), 2, 3, byrow = TRUE), 
  	widths=c(4,4,1), heights=c(1,1))

par(las=2)
par(mar=c(11,4,2,1)+0.1)
barplot(t(economy[1:10,c("TOTAL_PROPDMG","TOTAL_CROPDMG")]/1000000),
        main="Total Damage Value (historic, Millions $)",
        log="y",
        offset=0.0001, #avoids problems with log(0)
        names.arg=economy$EVTYPE[1:10],
        horiz=FALSE,
        cex.axis = 0.8)
barplot(t(economy_lasttenyears[1:10,c("TOTAL_PROPDMG","TOTAL_CROPDMG")]/1000000),
        main="Total Damage Value (2001-2011, Millions $)",
        log="y",
        offset=0.0001,
        names.arg=economy_lasttenyears$EVTYPE[1:10],
        horiz=FALSE,
        cex.axis = 0.8)

# setup for no margins on the legend
par(mar=c(0, 0, 0, 0))
plot.new()
legend('center','groups',c("Property","Crop"), fill = c("black","grey"),horiz = FALSE)

par(las=2)
par(mar=c(11,4,2,1)+0.1)
barplot(t(economy[1:10,c("MEAN_PROPDMG","MEAN_CROPDMG")]/1000000),
        main="Mean Damage Value (historic, Millions $)",
        log="y",
        offset=0.0001,
        names.arg=economy$EVTYPE[1:10],
        horiz=FALSE,
        cex.axis = 0.8)
barplot(t(economy_lasttenyears[1:10,c("MEAN_PROPDMG","MEAN_CROPDMG")]/1000000),
        main="Mean Damage Value (2001-2011, Millions $)",
        log="y",
        offset=0.0001,
        names.arg=economy_lasttenyears$EVTYPE[1:10],
        horiz=FALSE,
        cex.axis = 0.8)
```

It can be observed that the main damages reported are the ones concerning properties. It can be also observed, as in the case of public health, that some events such as floods, tornados, or hail, accumulate a high damage because of their frequency (high total damage value, relatively small mean damage value), whereas other events, such as hurricane/typhoon, or storm surges, cause severe damages every time they happen, although their frequency is low (they do not appear in the 20 most frequent weather events, but the total damage value is high).